---
title: "Sample Size Estimation using a Latent Variable Model for Mixed Outcome Co-Primary, Multiple Primary and Composite Endpoints: Empirical results"
author: "Martina McMenamin"
date: 2021-11-20
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source(file.path('R', 'load_packages.R'))
source(file.path('R', 'helper_functions.R'))
```

This file generates the results shown in the manuscript 'Sample Size Estimation using a Latent Variable Model for Mixed Outcome Co-Primary, Multiple Primary and Composite Endpoints' by McMenamin et al. Some of the empirical results related to the composite endpoint are not generated within this file as they required the model to be fitted to generated data at each step and therefore require large amounts of computational power. The code for the sample size and power for the composite endpoint are still shown for completeness and code for the R package and Shiny app are available at <https://github.com/martinamcm>. 

#### Power for MUSE example - Figure 1

```{r}

# Define inputs 

nest<-seq(0,800,10)

Sigcorr<-matrix(0,4,4)
Sigcorr[1,1]<-1
Sigcorr[2,2]<-1
Sigcorr[3,3]=Sigcorr[4,4]<- 1
Sigcorr[1,2]=Sigcorr[2,1]<- 0.448
Sigcorr[1,3]=Sigcorr[3,1]<- 0.521
Sigcorr[1,4]=Sigcorr[4,1]<- 0.003
Sigcorr[2,3]=Sigcorr[3,2]<- 0.445
Sigcorr[2,4]=Sigcorr[4,2]<- (-0.0309)
Sigcorr[3,4]=Sigcorr[4,3]<- 0.066

delta_est = c(0.88/sqrt(18), 0.38/sqrt(0.35), 0.24, 0.40)

# Co-primary
power_coprim <- 
  map(.x = nest, 
      .f = power.known.var,
      K = 4, 
      Sigma = Sigcorr, 
      delta = delta_est,
      power = NULL,
      sig.level = 0.05
)  %>%
  map_dbl(pluck, "power") 


#Multiple primary

power_mult <- 
    map(.x = nest, 
        .f = atleast.one.endpoint,
        K = 4, 
        Sigma = Sigcorr, 
        delta = delta_est,
        power = NULL,
        sig.level = 0.05/4
)  %>%
  map_dbl(pluck, "power") 


# Composite

power_comp <- 
  map_dbl(
    .x = nest,
    .f = powerfunc,
    mean = 0.1281464,
    var = 0.04801441,
    alpha = 0.05
)

# Individual

power_y1 <- 
    map_dbl(
      .x = nest,
      .f = powerfunc,
      mean = 0.88,
      var = 18,
      alpha = 0.05
)

power_y2 <-
    map_dbl(
      .x = nest,
      .f = powerfunc,
      mean = 0.38,
      var = 0.35,
      alpha = 0.05
)

power_y3 <-
    map_dbl(
      .x = nest,
      .f = powerfunc,
      mean = 0.24,
      var = 1,
      alpha = 0.05
)

power_y4 <-
    map_dbl(
      .x = nest,
      .f = powerfunc,
      mean = 0.40,
      var = 1,
      alpha = 0.05
)

data_power <- data.frame(power = c(power_coprim, power_mult, power_comp,
                                   power_y1, power_y2, power_y3, power_y4),
                         endpoint = c(rep("Co-primary", length(nest)),
                                      rep("Multiple Primary", length(nest)),
                                      rep("Composite", length(nest)),
                                      rep("SLEDAI", length(nest)),
                                      rep("PGA", length(nest)),
                                      rep("BILAG", length(nest)),
                                      rep("Taper", length(nest))
                                      ),
                        sample = rep(nest, 7) 
)

data_power$endpoint <- factor(data_power$endpoint,
                             levels=c(
                               "Co-primary","Multiple Primary","Composite","SLEDAI","PGA","BILAG","Taper"))

Powerplot <- 
  ggplot(data_power,
         aes(x = sample,
             y = power,
             color = endpoint)
         ) + 
  geom_line(
    aes(linetype = endpoint)
    ) +
  scale_linetype_manual(
    values = c(
      rep("solid",3),
      rep("dotdash",4)
      )
    ) +
  ylim(c(0,1)) + 
  geom_hline(
    yintercept = 0.8, linetype=3
    ) +
  ylab("Power") + 
  xlab("Sample size per group") + 
  scale_color_brewer(palette = "Dark2") + 
  theme_light()

Powerplot

```



#### Sample sizes for co-primary and multiple primary - Table 1

```{r}

delta1 <- c(0.88/sqrt(18), 0.38/sqrt(0.35), 0.24, 0.40)
delta2 <- c(0.88/sqrt(19), 0.38/sqrt(0.35), 0.24, 0.40)
delta3 <- c(0.88/sqrt(20), 0.38/sqrt(0.35), 0.24, 0.40)

delta4 <- c(0.88/sqrt(18), 0.38/sqrt(0.45), 0.24, 0.40)
delta5 <- c(0.88/sqrt(18), 0.38/sqrt(0.55), 0.24, 0.40)
delta6 <- c(0.88/sqrt(18), 0.38/sqrt(0.65), 0.24, 0.40)

delta <- list(delta1, delta2, delta3, delta4, delta5, delta6)

samplesize_coprim <-
    lapply(X = delta, 
       FUN = power.known.var,
       K = 4,
       n = NULL,
       Sigma = Sigcorr,
       sig.level = 0.025,
       power = 0.80
       ) 

samplesize_mult <-     
  lapply(X = delta, 
       FUN = atleast.one.endpoint,
       K = 4,
       n = NULL,
       Sigma = Sigcorr,
       sig.level = 0.025/4,
       power = 0.80
       ) 
  
# sample size in each arm for univariate endpoints - show power = 80%

  n_y1 <-  
    map2_dbl(.x = c(18, 19, 20),
            .y = c(365, 386, 406),
      .f = powerfunc, 
      mean = 0.88,
      alpha = 0.025)
    
  n_y2 <- 
    map2_dbl(
      .x = c(0.45, 0.55, 0.65),
      .y = c(49, 60, 71),
      .f = powerfunc, 
      mean = 0.38,
      alpha = 0.025)
    
  n_y3 <- powerfunc(0.24, 1, 0.025, 273)
  n_y4 <- powerfunc(0.40,1,0.025,99)

```


#### Overall power for different correlation based on MUSE example - Figure 2

```{r}

SigcorrH<-matrix(0,4,4)
SigcorrH[1,1] = SigcorrH[2,2] = SigcorrH[3,3]=SigcorrH[4,4] <- 1
SigcorrH[1,2]=SigcorrH[2,1] = SigcorrH[1,3] = SigcorrH[3,1] = SigcorrH[1,4]=SigcorrH[4,1] =
  SigcorrH[2,3]=SigcorrH[3,2] = SigcorrH[2,4]=SigcorrH[4,2] = SigcorrH[3,4]=SigcorrH[4,3] <- 0.8

Sigcorr0<-matrix(0,4,4)
Sigcorr0[1,1] = Sigcorr0[2,2] = Sigcorr0[3,3]=Sigcorr0[4,4]<-1
Sigcorr0[1,2]=Sigcorr0[2,1] = Sigcorr0[1,3]=Sigcorr0[3,1] = Sigcorr0[1,4]=Sigcorr0[4,1] = Sigcorr0[2,3]=Sigcorr0[3,2] = Sigcorr0[2,4]=Sigcorr0[4,2] = Sigcorr0[3,4]=Sigcorr0[4,3] <- 0

SigcorrL<-matrix(0,4,4)
SigcorrL[1,1] = SigcorrL[2,2] = SigcorrL[3,3]=SigcorrL[4,4] <- 1
SigcorrL[1,2]=SigcorrL[2,1] = SigcorrL[1,3]=SigcorrL[3,1] = SigcorrL[1,4]=SigcorrL[4,1] = SigcorrL[2,3]=SigcorrL[3,2] = SigcorrL[2,4]=SigcorrL[4,2] = SigcorrL[3,4]=SigcorrL[4,3]<-0.3


SigcorrM<-matrix(0,4,4)
SigcorrM[1,1] = SigcorrM[2,2] = SigcorrM[3,3]=SigcorrM[4,4]<-1
SigcorrM[1,2]=SigcorrM[2,1] = SigcorrM[1,3]=SigcorrM[3,1] = SigcorrM[1,4]=SigcorrM[4,1] = SigcorrM[2,3]=SigcorrM[3,2] = SigcorrM[2,4]=SigcorrM[4,2] = SigcorrM[3,4]=SigcorrM[4,3] <- 0.5


nest<-seq(0,500,10)

# co-primary
power_co_0 <- 
  map(
    .x = nest, 
    .f = power.known.var,
    K = 4,
    Sigma = Sigcorr0,
    delta = delta_est,
    sig.level = 0.05, 
    power = NULL
) %>% 
  map_dbl(pluck, "power") 
  
power_co_L <- 
  map(
    .x = nest, 
    .f = power.known.var,
    K = 4,
    Sigma = SigcorrL,
    delta = delta_est,
    sig.level = 0.05, 
    power = NULL
) %>% 
  map_dbl(pluck, "power") 

power_co_M <- 
  map(
    .x = nest, 
    .f = power.known.var,
    K = 4,
    Sigma = SigcorrM,
    delta = delta_est,
    sig.level = 0.05, 
    power = NULL
) %>% 
  map_dbl(pluck, "power") 


power_co_H <- 
  map(
    .x = nest, 
    .f = power.known.var,
    K = 4,
    Sigma = SigcorrH,
    delta = delta_est,
    sig.level = 0.05,
    power = NULL
) %>% 
  map_dbl(pluck, "power") 

power_corr_coprim <- c(power_co_0, power_co_L, power_co_M, power_co_H)

# Multiple 

nest_low <- seq(0,200,10)
  
power_mult_0 <- 
  map(
    .x = nest_low, 
    .f = atleast.one.endpoint,
    K = 4,
    Sigma = Sigcorr0,
    delta = delta_est,
    sig.level = 0.05/4, 
    power = NULL
) %>% 
  map_dbl(pluck, "power") 
  
power_mult_L <- 
  map(
    .x = nest_low, 
    .f = atleast.one.endpoint,
    K = 4,
    Sigma = SigcorrL,
    delta = delta_est,
   sig.level = 0.05/4, 
    power = NULL
) %>% 
  map_dbl(pluck, "power") 

power_mult_M <- 
  map(
    .x = nest_low, 
    .f = atleast.one.endpoint,
    K = 4,
    Sigma = SigcorrM,
    delta = delta_est,
    sig.level = 0.05/4, 
    power = NULL
) %>% 
  map_dbl(pluck, "power") 


power_mult_H <- 
  map(
    .x = nest_low, 
    .f = atleast.one.endpoint,
    K = 4,
    Sigma = SigcorrH,
    delta = delta_est,
    sig.level = 0.05/4, 
    power = NULL
) %>% 
  map_dbl(pluck, "power") 

power_corr_mult <- c(power_mult_0, power_mult_L, power_mult_M, power_mult_H)

# Composite 

  var0 = 0.08
  varL = 0.07
  varM = 0.06
  varH = 0.05
  
var_ests <- c(var0, varL, varM, varH)

power_comp_0 <- map_dbl(
  .x = nest_low,
  .f = powerfunc, 
  mean = 0.12,
  alpha = 0.05,
  var = var0
)

power_comp_L <- map_dbl(
  .x = nest_low,
  .f = powerfunc, 
  mean = 0.12,
  alpha = 0.05,
  var = varL
)

power_comp_M <- map_dbl(
  .x = nest_low,
  .f = powerfunc, 
  mean = 0.12,
  alpha = 0.05,
  var = varM
)

power_comp_H <- map_dbl(
  .x = nest_low,
  .f = powerfunc, 
  mean = 0.12,
  alpha = 0.05,
  var = varH
)

power_corr_comp <- c(power_comp_0, power_comp_L, power_comp_M, power_comp_H)


# Combine 

power_total <- c(power_corr_coprim, power_corr_mult, power_corr_comp)
corr_temp <- c(rep("Zero", length(power_co_0)),
              rep("Low", length(power_co_L)),
              rep("Medium", length(power_co_M)),
              rep("High", length(power_co_H))
              )
corr_temp_low <- c(rep("Zero", length(power_mult_0)),
              rep("Low", length(power_mult_L)),
              rep("Medium", length(power_mult_M)),
              rep("High", length(power_mult_H))
              )
Correlation <- c(corr_temp, rep(corr_temp_low, 2))
endpoints <- c(rep("Co-primary", length(power_corr_coprim)),
               rep("Multiple Primary", length(power_corr_mult)),
               rep("Composite", length(power_corr_comp))
               )

sample = c(rep(nest, 4) , rep(nest_low, 8))

data_power_corr <- data.frame(sample, power_total, Correlation, endpoints)

Powercorr <- 
  ggplot(data_power_corr,
         aes(x = sample,
             y = power_total,
             color = Correlation)) + 
  geom_line() + 
  ylim(c(0,1)) + 
  geom_hline(yintercept = 0.8,linetype=3) +
  ylab("Power") + 
  xlab("Sample size per group") + 
  scale_color_viridis(discrete = "TRUE") + 
  theme_bw() +
  facet_wrap(~endpoints, scales="free")+
  labs(color="Correlation")

```



#### MUSE composite sample size - Table 2

```{r}

mean.lat = 0.20
sig.lat = c(0.05, 0.06, 0.07, 0.08, 0.09, 0.10)
alpha = 0.05 # two sided 0.10 in MUSE trial
beta = 0.12 # 88% power required in MUSE trial

ests = sapply(X = sig.lat, 
       FUN = samplesize,
         mean.lat = 0.20, 
         mean.bin = 0.20, 
         var.bin = 0.16, # derived using standard p*(1-p)
         alpha = alpha, 
         beta = beta
       )

total_sample_size = ceiling(ests[1,])*2

# Empirical power not shown can't run the model so use variation in 
# estimates of sigma to estimate directly - detailed properties of 
# latent variable method for composite endpoints shown in 
# McMenamin M, Barrett JK, Berglind A, Wason JM. Employing a latent variable framework to improve efficiency in composite endpoint analysis. Statistical Methods in Medical Research. 2021;30(3):702-716. doi:10.1177/0962280220970986

```


#### Empirical power - Table 3

```{r}

#### TABLE 3 

# First obtain tables of sample sizes for K = 4 outcomes with high, medium, low and zero correlation between endpoints
 
K = 4
var_1 = var_2 = var_3 = var_4 <- 1
rho_H <- 0.8
rho_M <- 0.5
rho_L <- 0.3
rho_0 <- 0
delta_1 = rep(0.12, 4)
delta_2 = c(rep(0.35, 2), rep(0.15, 2))
delta_3 = c(0.12, 0.35, 0.55, 0.10)
delta = list(delta_1, delta_2, delta_3)

Sigma_H = Sigma_M = Sigma_L = Sigma_0 <- matrix(0, nrow = K, ncol = K)
diag(Sigma_H) = diag(Sigma_M) = diag(Sigma_L) = diag(Sigma_0) <- var_1

Sigma_H[Sigma_H != diag(Sigma_H)] = rho_H
Sigma_M[Sigma_M != diag(Sigma_M)] = rho_M
Sigma_L[Sigma_L != diag(Sigma_L)] = rho_L

Sigma_cor = list(Sigma_0, Sigma_L, Sigma_M, Sigma_H)


 # Co-primary

sample_co_1 <- 
  map(.x = Sigma_cor,
      .f = power.unknown.var, 
       K = K, delta = delta_1, sig.level = 0.025, power = 0.8, n = NULL) %>%
  map_dbl(pluck, "n")

sample_co_2 <- 
  map(.x = Sigma_cor,
      .f = power.known.var, 
       K = K, delta = delta_2, sig.level = 0.025, power = 0.8, n = NULL) %>%
  map_dbl(pluck, "n")

sample_co_3 <- 
  map(.x = Sigma_cor,
      .f = power.known.var, 
       K = K, delta = delta_3, sig.level = 0.025, power = 0.8, n = NULL) %>%
  map_dbl(pluck, "n") 

sample_co <- ceiling(c(sample_co_1, sample_co_2, sample_co_3))


  #### Multiple primary

# Apply Bonferroni correction for K endpoints 

sample_mult_1 <- 
  map(.x = Sigma_cor,
      .f = atleast.one.endpoint, 
       K = K, delta = delta_1, sig.level = 0.025/K, power = 0.8, n = NULL) %>%
  map_dbl(pluck, "n")

sample_mult_2 <- 
  map(.x = Sigma_cor,
      .f = atleast.one.endpoint, 
       K = K, delta = delta_2, sig.level = 0.025/K, power = 0.8, n = NULL) %>%
  map_dbl(pluck, "n")

sample_mult_3 <- 
  map(.x = Sigma_cor,
      .f = atleast.one.endpoint, 
       K = K, delta = delta_3, sig.level = 0.025/K, power = 0.8, n = NULL) %>%
  map_dbl(pluck, "n") 

sample_mult <- ceiling(c(sample_mult_1, sample_mult_2, sample_mult_3))


  # Composite 

# Note this file doesn't run the composite results as these need a lot of 
# computing power and to be run in parallel as the model needs fit to multiple
# datasets to estimate variance in order to obtain sample size - functions to obtain 
# power and sample size from estimated quantities included in source file for completeness 





# Obtain empirical power for each sample size shown

  # Co-primary

set.seed(37889)

pval = vector()
emp_power = vector()
delta_val = vector()
sig_val = list()
nsim = 1000 # NOTE! increase this to 100000 to get values in manuscript

for(j in 1:length(sample_co)){

  samp_size = sample_co[j]
  delta_val = ifelse(j<5, delta[1], 
                     ifelse(j>4 && j<9, delta[2], delta[3]))
  sig_val = ifelse(j %in% c(1,5,9), Sigma_cor[1], ifelse(j %in% c(2,6,10), Sigma_cor[2], ifelse( j %in% c(3,7,11), Sigma_cor[3], Sigma_cor[4])))
  
for(i in 1:nsim){
  
# Set Y as control group 
X = rmvnorm(n=samp_size, mean = unlist(delta_val), sigma = sig_val[[1]])
Y = rmvnorm(n=samp_size, mean = rep(0, length(unlist(delta_val))), sigma = sig_val[[1]])

pval[i] <- ifelse(mpe.z.test(X, Y, Sigma=sig_val[[1]])$p.value < 0.025, 1, 0)

}

  emp_power[j] = (sum(pval)/nsim)*100
  
}

# Table of sample sizes and empirical powers for co-primary endpoints in Table 3
emp_power_co <- tibble(sample_co, emp_power)



  # Multiple primary

# Define function for union-intersection z test

mult.z.test <- function(X, Y, Sigma, conf.level = 0.975){
  
   diff <- colMeans(X) - colMeans(Y)
    names(diff) <- paste("EP", 1:ncol(X), sep = ".")
    n <- nrow(X)
    m <- nrow(Y)
    SD <- sqrt(diag(Sigma))
    Z <- diff/(SD * sqrt(1/n + 1/m))
    
    statistic <- max(Z)
    names(statistic) <- "z"
    pval <- min(pnorm(Z, lower.tail = FALSE))
    alpha <- (1 - conf.level)/4
    cint <- rep(qnorm(1 - alpha), ncol(X))
    cint <- cbind(Z - cint, Inf) * SD * sqrt(1/n + 1/m)
    colnames(cint) <- c(paste((1 - conf.level)/4), 1)
    rownames(cint) <- paste("EP", 1:ncol(X), sep = ".")
    
    alternative <- "true difference in means is larger than 0 for at least one endpoint"
    attr(cint, "conf.level") <- conf.level
    method <- "Union-intersection z-test"
    rval <- list(method = method, statistic = statistic, p.value = pval, 
        conf.int = cint, estimate = diff, alternative = alternative)
    class(rval) <- "mpe.test"
    return(rval)
  
}

set.seed(37889)

pval = vector()
emp_power = vector()
delta_val = vector()
sig_val = list()
nsim = 1000 # NOTE! increase this to 100000 to get values in manuscript

for(j in 1:length(sample_mult)){

  samp_size = sample_mult[j]
  delta_val = ifelse(j<5, delta[1], 
                     ifelse(j>4 && j<9, delta[2], delta[3]))
  sig_val = ifelse(j %in% c(1,5,9), Sigma_cor[1], ifelse(j %in% c(2,6,10), Sigma_cor[2], ifelse( j %in% c(3,7,11), Sigma_cor[3], Sigma_cor[4])))
  
for(i in 1:nsim){
  
# Set Y as control group 
X = rmvnorm(n=samp_size, mean = unlist(delta_val), sigma = sig_val[[1]])
Y = rmvnorm(n=samp_size, mean = rep(0, length(unlist(delta_val))), sigma = sig_val[[1]])

pval[i] <- ifelse(mult.z.test(X, Y, Sigma=sig_val[[1]])$p.value < 0.025/4, 1, 0)

}

  emp_power[j] = (sum(pval)/nsim)*100
  
}

emp_power


# Table of sample sizes and empirical powers for multiple primary endpoints in Table 3
emp_power_mult <- tibble(sample_mult, emp_power)

```



#### Family-wise error rate - Figure 3


```{r}

### FIGURE 3


# Show effect of different correlations between endpoints on both unadjusted and adjusted FWER 

# Add more Sigma cols 

Sigma_VH <- matrix(0, nrow = K, ncol = K)
diag(Sigma_VH) <- var_1
Sigma_VH[Sigma_VH != diag(Sigma_VH)] = 1

Sigma_FWER_4 <- list(Sigma_0, Sigma_L, Sigma_M, Sigma_H, Sigma_VH)
Sigma_FWER_3 <- list(Sigma_0[1:3, 1:3], Sigma_L[1:3, 1:3], Sigma_M[1:3, 1:3], 
                     Sigma_H[1:3, 1:3], Sigma_VH[1:3, 1:3])
Sigma_FWER_2 <- list(Sigma_0[1:2, 1:2], Sigma_L[1:2, 1:2], Sigma_M[1:2, 1:2], 
                     Sigma_H[1:2, 1:2], Sigma_VH[1:2, 1:2])

# 4 endpoints 

FWER_unadj_4 <- map(.x = Sigma_FWER_4,
                  .f = FWER,
                  K=4, delta = c(0,0,0,0), n=300, adjust = FALSE) %>%
  map_dbl(pluck, "FWER")


FWER_adj_4 <- map(.x = Sigma_FWER_4,
                  .f = FWER,
                  K=4, delta = c(0,0,0,0), n=300, adjust = TRUE) %>%
  map_dbl(pluck, "FWER")


# 3 endpoints

FWER_unadj_3 <- map(.x = Sigma_FWER_3,
                  .f = FWER,
                  K=3, delta = c(0,0,0), n=300, adjust = FALSE) %>%
  map_dbl(pluck, "FWER")


FWER_adj_3 <- map(.x = Sigma_FWER_3,
                  .f = FWER,
                  K=3, delta = c(0,0,0), n=300, adjust = TRUE) %>%
  map_dbl(pluck, "FWER")


# 2 endpoints

FWER_unadj_2 <- map(.x = Sigma_FWER_2,
                  .f = FWER,
                  K=2, delta = c(0,0), n=300, adjust = FALSE) %>%
  map_dbl(pluck, "FWER")


FWER_adj_2 <- map(.x = Sigma_FWER_2,
                  .f = FWER,
                  K=2, delta = c(0,0), n=300, adjust = TRUE) %>%
  map_dbl(pluck, "FWER")


# Overall results

FWER_results <- tibble(corr = rep(c(0, 0.3, 0.5, 0.8, 1.0), 3), 
                       endpt = c(rep("4", 5), rep("3", 5), rep("2", 5)),
                       FWER_unadj = c(FWER_unadj_4, FWER_unadj_3, FWER_unadj_2),
                       FWER_adj = c(FWER_adj_4, FWER_adj_3, FWER_adj_2)
                       )

ggplot(data = FWER_results) +
  geom_line(aes(x = corr, y = FWER_unadj, linetype = endpt,  colour = "Unadjusted")) +
  geom_line(aes(x = corr, y = FWER_adj, linetype = endpt, colour = "Adjusted")) +
  geom_point(aes(x = corr, y = FWER_adj), shape = 4, size = 2, colour = "DarkBlue") +
  geom_point(aes(x = corr, y = FWER_unadj), shape = 4, size = 2, colour = "DarkGreen") +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank())+
  geom_hline(yintercept = 0.025) +
  ylim(c(0,0.10)) +
  ylab("Family wise error rate") +
  xlab("Correlation") +
  labs(linetype = "Number of \noutcomes",
       colour = "Type") +
  scale_colour_manual(values = c("DarkBlue", "DarkGreen"))
  


```
